Geometric signs reveal changes social structures and networks during the Western Eurasian Aurignacian

### 2025-06-09 paper/paper.qmd
edits to text, some plots, formatting, front matter

---
title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
author: Sophie Cooper
format: docx
echo: false
execute: 
  warning: false
bibliography: references.bib
link-citations: true
---
# Introduction

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans. Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of objects to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian. 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018]. Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded. Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters. To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks. Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.

In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. The Aurignacian can be divided into phases which are mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @tbl-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.

@@

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using perMANOVA tests. All of our programming in this study was conducted in RStudio.

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

@@

The distribution of objects across sites is highly skewed, as seen in @fig-site-map C. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]. The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

If the distribution of sign types can be explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015]. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types (1-3 types), which tend to be simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs. Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.

In the Transitional phase, there is a relatively small range of sign types compared to other phases. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.

In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader group.

In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line). The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. In addition to line and notch, the obline, dot and vulva are also frequently found in the broader group.

Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader group.

@@

The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provides additional insights into the evolution of connectivity between groups. In the Transitional phase, the network analysis reveals one distinct and one disconnected group. The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consists of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups. The Proto-Aurignacian phase also shows two clearly defined groups, the broad group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud. However, it also displays some connectivity between the two groups of sites, with La Ferrassie connected to both groups. Les Cottés stands out with no connections to either group. The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups. Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois. As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.

```{r fig.height= 7, fig.width=10}
#| label: fig-group-map
#| fig-cap: "Map of sites faceted by time period and colored by group number"

ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf, aes(color = group)) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2,
                  bg.color = "white",
                  bg.r = 0.25) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        strip.text = element_text(size = 10))

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

@@

The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership. The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`. We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods. The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`. Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period. The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).

@@

## Grouping Dynamics Across the Four Phases of the Aurignacian
Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

>>>>

---
title: "Geometric signs reveal changes social structures and networks during the Western Eurasian Aurignacian"
author:
  - Sophie Cooper:
      institute: uw
      orcid: xxx
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - uw: Department of Anthropology, University of Washington, Seattle, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  ...
keywords: |

---

# Introduction

Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans. Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is often linked to the dispersal of _Homo sapiens_ across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper we investigate Aurignacian social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.

Previous work has fruitfully explored social organization dynamics of the Upper Paleolithic using statistical analyses of objects to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicate ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian period. They used seriation, correspondence, and GIS analyses of 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East to identify geographically cohesive groups sharing similar ornament type associations. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

Cultural groups have also been identified from forty-two quadrilateral geometric signs from four cave sites in the Cantabrian region of Spain, likely associated with the Magdalenian period, spanning a distance of 30 kilometers [@Sauvet_2018]. Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters. To interpret these clusters, @Sauvet_2018 drew on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks. Using these case studies as analogies, @Sauvet_2018 interpreted the Magdalenian geometric signs as identity markers, and the two clusters as representative of two cultural groups.

In this study, we extend this previous work using geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43-30 k BP, and is largely characterized by material culture indicators such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early, 43–37 k BP), and Aur-P2 (Evolved/Late, 37-32 k BP). @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8. Similar to @Banks_dErrico_Zilhao_2013, we divide the Aurignacian into four phases based on material culture changes; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as summarised in @tbl-aur-table.

@@

We hypothesize that these phases, and the environmental changes across the them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? What do these patterns tell us about dynamics of social organization throughout the Aurignacian? To answer these questions we use seriation analysis, network analysis, and PerMANOVA tests.

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs classified by @dutkiewicz2020 as "other". To improve the quality of our data we excluded seven sites from the sample: only Willendorf (one sign present); Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (extreme geographic distance from the other sites); Grotte De La Princesse Pauline and Šandalja II (dated much later than the rest of the assemblages). This resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

@@

The distribution of objects across sites is highly skewed, as seen in @fig-site-map C. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value estimates the probability of obtaining the observed correlation if the null hypothesis of no correlation were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]. The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`), validating our decision to use presence/absence data.

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

If the distribution of sign types can be mostly explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015]. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so we reject the isolation-by-distance hypothesis as a primary explanation for variation in sign type distribution.

@@

For each of the phases, our seriation analysis identifies two groups, one with a restricted range of sign types (1-3 types), which tend to be visually simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs. Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods. Overall, across the four phases, the number of sites in the restricted range group tends to be less than or equal to the number of sites in the broader range group, with the exception of the final Evolved phase.

In the Transitional phase, there is a relatively small range of sign types compared to other phases. The restricted range group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader range group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.

In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted range group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline. The broader range group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader range group of the Transitional phase. In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader range group.

In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted range group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line). The broader range group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. In addition to line and notch, the obline, dot and vulva are also frequently found in the broader range group.

Finally, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian. The restricted range group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs. The broader range group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader range group.

@@

The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provide additional insights into the evolution of connectivity between groups. In the Transitional phase, the network analysis reveals one distinct and one disconnected group. The restricted range group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad range group consists of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups. The Proto-Aurignacian phase also shows two clearly defined groups, the broad range group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted range group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud. However, it also displays some connectivity between the two groups of sites, with La Ferrassie connected to both groups. Les Cottés stands out with no connections to either group. The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups. Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad range group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted range group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois. As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.

```{r fig.height= 7, fig.width=10}
#| label: fig-group-map
#| fig-cap: "Map of sites faceted by time period and colored by group type"

signbase_sf_groups <- 
  signbase_sf %>% 
  mutate(group_map = case_when(
    group == 1 ~ "Restricted range",
    group == 2 ~ "Broader range",
  ))

ggplot(Europe) +
  geom_sf(fill = "white") +
  geom_sf(data = signbase_sf_groups, 
          aes(shape = group_map),
          color = "black",
          size = 3.5) +
  geom_sf(data = signbase_sf_groups, 
          aes(color = group_map,
              shape = group_map),
          size = 3) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf_groups,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 3,
                  bg.color = "white",
                  bg.r = 0.1) +
  theme_bw() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        strip.text = element_text(size = 10),
        legend.position = "bottom")  +
  scale_color_manual(name = "Group", 
                     values = c("Restricted range" = "#FFFFB3", 
                                "Broader range"    = "#8DD3C7")) +
  scale_shape_manual(name = "Group",
                     values = c("Restricted range" = 17, 
                                "Broader range"    = 19))

@@

While th results of the seriation and network analyses point to two main groups of sites within each phase, spatial clusters of sites are not highly evident in these groups, in part due to the small numbers of sites in each phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower geographic distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

@@

The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership. The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`. We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods. The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`. Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period. The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad range groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).

@@

## Grouping Dynamics Across the Four Phases of the Aurignacian

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type number and diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to very strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

One consistent grouping dynamic throughout all four phases is the restricted/broader range group divide, which may be influenced by variation in social structure. Variation in hunter-gatherer social structures is well-known from ethnographic and archaeological research  [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012]. Studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about the numbers and types of social roles and statuses within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012]. Drawing on this work, we interpret the broader/restricted range grouping in the geometric sign data as indicating variation in social structure. The broader range groups, with higher variety of signs and more visually complex signs, indicates communities with a greater variety of formalized social roles. Conversely, the sites with a restricted range of signs indicate communities with fewer formalized social roles. A single larger community could have both of these groups as sub-sets, rather than the restricted/broader range groups being distinct, non-overlapping cultural groups.
